---
title: "Lab 5"
author: "Kendall Dimson"
format: pdf
editor: visual
embed-resources: true
fig-width: 3
fig-height: 3
---

## Set-Up

```{r}
install.packages("data.table")
library(data.table)
library(tidyverse)
library(dtplyr)

# Download the data
stations <- fread("https://noaa-isd-pds.s3.amazonaws.com/isd-history.csv")
stations[, USAF := as.integer(USAF)]

# Dealing with NAs and 999999
stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

# Selecting the three relevant columns, and keeping unique records
stations <- unique(stations[, list(USAF, CTRY, STATE)])

# Dropping NAs
stations <- stations[!is.na(USAF)]

# Removing duplicates
stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]

if (!file.exists("met_all.gz"))
  download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method   = "libcurl",
    timeout  = 60
    )
met <- data.table::fread("met_all.gz")

#merge datasets
dat <- merge(
  # Data
  x     = met,      
  y     = stations, 
  # List of variables to match
  by.x  = "USAFID",
  by.y  = "USAF", 
  # Which obs to keep?
  all.x = TRUE,      
  all.y = FALSE
  )
head(dat[, c('USAFID', 'WBAN', 'STATE')], n = 4)

dat <- as.data.frame(dat)

```

## Question 1

What is the median station in terms of temperature, wind speed, and atmospheric pressure? Look for the three weather stations that best represent continental US using the quantile() function. Do these three coincide?

```{r}
median_temp <- quantile(dat$temp, na.rm=TRUE, 0.5)
median_wind.sp <- quantile (dat$wind.sp,  na.rm=TRUE,0.5)
median_atm.press <- quantile(dat$atm.press, na.rm=TRUE, 0.5)

station_temp<- dat |> filter (temp == median_temp) |> select('USAFID')
station_wind.sp<- dat |> filter (wind.sp == median_wind.sp) |> select('USAFID')
station_atm.press<- dat |> filter (atm.press == median_atm.press) |> select('USAFID')

print(station_temp, station_wind.sp, station_atm.press)
```
Stations 690150 and 720113 coincide.

## Question 2

Just like the previous question, you are asked to identify what is the most representative, the median, station per state. This time, instead of looking at one variable at a time, look at the euclidean distance. If multiple stations show in the median, select the one located at the lowest latitude.

```{r}
medians<- dat |> summarise(
    median_temp = median(temp),
    median_wind.sp = median(wind.sp),
    median_atm.press = median(atm.press)
)

most_representative_stations <- dat|>
  group_by(STATE) |>
  filter(distance == min(distance)) %>%
  slice(which.min(lat)) %>%  # If multiple, select the one with the lowest latitude
  ungroup()
```


## Question 3

For each state, identify what is the station that is closest to the mid-point of the state. Combining these with the stations you identified in the previous question, use leaflet() to visualize all ~100 points in the same figure, applying different colors for those identified in this question.

Knit the doc and save it on GitHub.

## Question 4

Using the quantile() function, generate a summary table that shows the number of states included, average temperature, wind-speed, and atmospheric pressure by the variable “average temperature level,” which you’ll need to create.

Start by computing the states' average temperature. Use that measurement to classify them according to the following criteria:

low: temp < 20
Mid: temp >= 20 and temp < 25
High: temp >= 25
Once you are done with that, you can compute the following:

Number of entries (records),
Number of NA entries,
Number of stations,
Number of states included, and
Mean temperature, wind-speed, and atmospheric pressure.
All by the levels described before.

Knit the document, commit your changes, and push them to GitHub. Once you’re done, you can email me the link to your repository or, preferably, link to the Lab 5 Issue in your commit message, as follows:

git commit -a -m "Finalizing lab 5 https://github.com/USCbiostats/PM566/issues/68"
This will let me know which version of your repository to look at for grading purposes.
